Efficient High-Level Abstractions for Web Programming

### 2014-03-31 papers/gdrgpl2014/gdr-foy.tex
initial version

\documentclass{sigplanconf}

@@

\conferenceinfo{GPCE~'13}{October 27--28, 2013, Indianapolis, Indiana, USA} 
\copyrightyear{2013} \copyrightdata{978-1-4503-2373-4/13/10} \doi{2517208.2517227}

@@

\authorinfo{Julien Richard-Foy\and Olivier Barais\and Jean-Marc Jézéquel}
           {IRISA, Université de Rennes 1}
           {\{first\}.\{last\}@irisa.fr}

@@

\category{D.3.3}{Programming Languages}{Language Constructs and Features}

@@

On one hand, for engineering reasons, developers want to write Web applications using a single
high-level language, abstracting over the target platforms differences and reducing code complexity.
But on the other hand, for performance reasons, they want to keep control on the way their code is
compiled to each target platform. We propose to solve this dilemma by providing high-level
abstractions in compiled domain-specific embedded languages (DSELs)~\cite{Hudak96_DSEL,
Elliott2003_Compiling}. Compiled DSELs allow the definition of domain-specific languages (DSLs) as
libraries on top of a host language, and to compile them to a target platform. Their deep embedding
gives the opportunity to control the code generation scheme for a given abstraction and target
platform.

@@

Lightweight Modular Staging~\cite{Rompf12_LMSThesis, Rompf12_LMS} (LMS) is a framework for defining
deeply embedded DSLs in Scala. It has been used to define high-performance DSLs for parallel
computing~\cite{Brown11_Parallel} and to define JavaScript as an embedded DSL in
Scala~\cite{Kossakowski12_JsDESL}.

@@

@@

In a Web application, the user interface is defined by a HTML document that can be updated by the
JavaScript code. A typical operation consists in searching some “interesting” element in the
document, in order to extract its content, replace it or listen to user events triggered on it (such
as mouse clicks). The standard API provides several functions to search elements in a HTML document
according to their name or attribute values. Figure~\ref{selectors-api} summarizes the available
functions and their differences.

@@

\bibliographystyle{abbrvnat}
\begin{thebibliography}{24}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi: #1}\else
  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[Ashkenas(2011)]{Ashkenas2011_CoffeeScript}
J.~Ashkenas.
\newblock Coffeescript, 2011.

\bibitem[Bibeault and Kats(2008)]{Bibeault08_jQuery}
B.~Bibeault and Y.~Kats.
\newblock \emph{jQuery in Action}.
\newblock Dreamtech Press, 2008.

\bibitem[Brown et~al.(2011)Brown, Sujeeth, Lee, Rompf, Chafi, Odersky, and
  Olukotun]{Brown11_Parallel}
K.~J. Brown, A.~K. Sujeeth, H.~J. Lee, T.~Rompf, H.~Chafi, M.~Odersky, and
  K.~Olukotun.
\newblock A heterogeneous parallel framework for domain-specific languages.
\newblock In \emph{Parallel Architectures and Compilation Techniques (PACT),
  2011 International Conference on}, pages 89--100. IEEE, 2011.

\bibitem[Cannasse(2008)]{Cannasse08_HaXe}
N.~Cannasse.
\newblock Using haxe.
\newblock \emph{The Essential Guide to Open Source Flash Development}, pages
  227--244, 2008.

\bibitem[Chaganti(2007)]{Chaganti07_GWT}
P.~Chaganti.
\newblock \emph{Google Web Toolkit: GWT Java Ajax Programming}.
\newblock Packt Pub Limited, 2007.

\bibitem[Cooper et~al.(2007)Cooper, Lindley, Wadler, and
  Yallop]{Cooper07_Links}
E.~Cooper, S.~Lindley, P.~Wadler, and J.~Yallop.
\newblock Links: Web programming without tiers.
\newblock In \emph{Formal Methods for Components and Objects}, pages 266--296.
  Springer, 2007.

\bibitem[Elliott et~al.(2003)Elliott, Finne, and
  De~Moor]{Elliott2003_Compiling}
C.~Elliott, S.~Finne, and O.~De~Moor.
\newblock Compiling embedded languages.
\newblock \emph{Journal of Functional Programming}, 13\penalty0 (3):\penalty0
  455--481, 2003.

\bibitem[Griffith(2011)]{Griffith11_Dart}
R.~Griffith.
\newblock The dart programming language for non-programmers-overview.
\newblock 2011.

\bibitem[Hoare(2009)]{Hoare09_Null}
T.~Hoare.
\newblock Null references: The billion dollar mistake.
\newblock \emph{Presentation at QCon London}, 2009.

\bibitem[Huang et~al.(2010)Huang, Xu, Tiwana, Mao, Zhang, and
  Bahl]{Huang10_Perf}
J.~Huang, Q.~Xu, B.~Tiwana, Z.~M. Mao, M.~Zhang, and P.~Bahl.
\newblock Anatomizing application performance differences on smartphones.
\newblock In \emph{Proceedings of the 8th international conference on Mobile
  systems, applications, and services}, pages 165--178. ACM, 2010.

\bibitem[Hudak(1996)]{Hudak96_DSEL}
P.~Hudak.
\newblock Building domain-specific embedded languages.
\newblock \emph{ACM Computing Surveys}, 28, 1996.
\newblock URL \url{http://dsel.ps}.

\bibitem[J{\o}rring and Scherlis(1986)]{Jorring1986_Staging}
U.~J{\o}rring and W.~L. Scherlis.
\newblock Compilers and staging transformations.
\newblock In \emph{Proceedings of the 13th ACM SIGACT-SIGPLAN symposium on
  Principles of programming languages}, pages 86--96. ACM, 1986.

\bibitem[Kossakowski et~al.(2012)Kossakowski, Amin, Rompf, and
  Odersky]{Kossakowski12_JsDESL}
G.~Kossakowski, N.~Amin, T.~Rompf, and M.~Odersky.
\newblock Java{S}cript as an {E}mbedded {DSL}.
\newblock In J.~Noble, editor, \emph{{ECOOP} 2012 -- {O}bject-{O}riented
  {P}rogramming}, volume 7313 of \emph{Lecture Notes in Computer Science},
  pages 409--434, Berlin, Heidelberg, 2012. Springer Berlin Heidelberg.
\newblock \doi{10.1007/978-3-642-31057-7_19}.
\newblock URL \url{https://github.com/js-scala/js-scala/}.

\bibitem[Kuuskeri and Mikkonen(2009)]{Kuuskeri09_PartitioningClientServer}
J.~Kuuskeri and T.~Mikkonen.
\newblock Partitioning web applications between the server and the client.
\newblock In \emph{Proceedings of the 2009 ACM symposium on Applied Computing},
  SAC '09, pages 647--652, New York, NY, USA, 2009. ACM.
\newblock ISBN 978-1-60558-166-8.
\newblock \doi{10.1145/1529282.1529416}.
\newblock URL \url{http://doi.acm.org/10.1145/1529282.1529416}.

\bibitem[McGranaghan(2011)]{McGranaghan11_ClojureScript}
M.~McGranaghan.
\newblock Clojurescript: Functional programming for javascript platforms.
\newblock \emph{Internet Computing, IEEE}, 15\penalty0 (6):\penalty0 97--102,
  2011.

\bibitem[Mikkonen and Taivalsaari(2008)]{Mikkonen08_SpaghettiJs}
T.~Mikkonen and A.~Taivalsaari.
\newblock Web applications - spaghetti code for the 21st century.
\newblock In \emph{Proceedings of the 2008 Sixth International Conference on
  Software Engineering Research, Management and Applications}, pages 319--328,
  Washington, DC, USA, 2008. IEEE Computer Society.
\newblock ISBN 978-0-7695-3302-5.
\newblock \doi{10.1109/SERA.2008.16}.
\newblock URL \url{http://dl.acm.org/citation.cfm?id=1443226.1444030}.

\bibitem[Nanda and Sinha(2009)]{Nanda09_Null}
M.~Nanda and S.~Sinha.
\newblock Accurate interprocedural null-dereference analysis for java.
\newblock In \emph{Software Engineering, 2009. ICSE 2009. IEEE 31st
  International Conference on}, pages 133--143. IEEE, 2009.

\bibitem[Preciado et~al.(2005)Preciado, Trigueros, S{\'a}nchez-Figueroa, and
  Comai]{Preciado05_RIAMethodologyNecessity}
J.~C. Preciado, M.~L. Trigueros, F.~S{\'a}nchez-Figueroa, and S.~Comai.
\newblock Necessity of methodologies to model rich internet applications.
\newblock In \emph{WSE}, pages 7--13. IEEE Computer Society, 2005.
\newblock ISBN 0-7695-2470-2.

\bibitem[Rodr{\'\i}guez-Echeverr{\'\i}a(2009)]{Echeverria09_RIA}
R.~Rodr{\'\i}guez-Echeverr{\'\i}a.
\newblock Ria: more than a nice face.
\newblock In \emph{Proceedings of the Doctolral Consortium of the International
  Conference on Web Engineering}, volume 484. CEUR-WS.org, 2009.

\bibitem[Rompf(2012)]{Rompf12_LMSThesis}
T.~Rompf.
\newblock \emph{Lightweight Modular Staging and Embedded Compilers: Abstraction
  without Regret for High-Level High-Performance Programming}.
\newblock PhD thesis, {\'E}COLE POLYTECHNIQUE F{\'E}D{\'E}RALE DE LAUSANNE,
  2012.

%\vfill\eject

\bibitem[Rompf et~al.(2012)Rompf, Sujeeth, Amin, Brown, Jovanovic, Lee,
  Jonnalagedda, Olukotun, and Odersky]{Rompf12_LMS}
T.~Rompf, A.~Sujeeth, N.~Amin, K.~Brown, V.~Jovanovic, H.~Lee, M.~Jonnalagedda,
  K.~Olukotun, and M.~Odersky.
\newblock Optimizing {D}ata {S}tructures in {H}igh-{L}evel {P}rograms: {N}ew
  {D}irections for {E}xtensible {C}ompilers based on {S}taging.
\newblock Technical report, 2012.

\bibitem[Souders(2008)]{Souders08_Perf}
S.~Souders.
\newblock High-performance web sites.
\newblock \emph{Communications of the ACM}, 51\penalty0 (12):\penalty0 36--41,
  2008.

\bibitem[Van~Deursen et~al.(2000)Van~Deursen, Klint, and Visser]{Van00_DSL}
A.~Van~Deursen, P.~Klint, and J.~Visser.
\newblock Domain-specific languages: an annotated bibliography.
\newblock \emph{ACM Sigplan Notices}, 35\penalty0 (6):\penalty0 26--36, 2000.

\bibitem[Visser(2007)]{Visser07}
E.~Visser.
\newblock {WebDSL}: {A} case study in domain-specific language engineering.
\newblock In R.~Lämmel, J.~Visser, and J.~Saraiva, editors, \emph{Generative
  and Transformational Techniques in Software Engineering II, International
  Summer School, {GTTSE} 2007}, volume 5235 of \emph{Lecture Notes in Computer
  Science}, pages 291--373, Braga, Portugal, 2007. Springer.
\newblock ISBN 978-3-540-88642-6.
\newblock \doi{http://dx.doi.org/10.1007/978-3-540-88643-3_7}.

\end{thebibliography}
%\begin{thebibliography}{}
%\softraggedright
%
%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%P. Q. Smith, and X. Y. Jones. ...reference text...
%
%\end{thebibliography}
%
\end{document}

>>>>

\documentclass{llncs}

@@

%\conferenceinfo{GPCE~'13}{October 27--28, 2013, Indianapolis, Indiana, USA} 
%\copyrightyear{2013} \copyrightdata{978-1-4503-2373-4/13/10} \doi{2517208.2517227}

@@

\author{Richard-Foy\and Olivier Barais\and Jean-Marc Jézéquel}

\institute{ IRISA, Université de Rennes 1
           {\email{\{first\}.\{last\}@irisa.fr}}}

@@

%\category{D.3.3}{Programming Languages}{Language Constructs and Features}

@@

On one hand, for engineering reasons, developers want to write Web applications using a single
high-level language, abstracting over the target platforms differences and reducing code complexity.
But on the other hand, for performance reasons, they want to keep control on the way their code is
compiled to each target platform. We propose to solve this dilemma by providing high-level
abstractions in compiled domain-specific embedded languages (DSELs)~\cite{Hudak96_DSEL,Elliott2003_Compiling}. Compiled DSELs allow the definition of domain-specific languages (DSLs) as
libraries on top of a host language, and to compile them to a target platform. Their deep embedding
gives the opportunity to control the code generation scheme for a given abstraction and target
platform.

@@

Lightweight Modular Staging~\cite{Rompf12_LMSThesis,Rompf12_LMS} (LMS) is a framework for defining
deeply embedded DSLs in Scala. It has been used to define high-performance DSLs for parallel
computing~\cite{Brown11_Parallel} and to define JavaScript as an embedded DSL in
Scala~\cite{Kossakowski12_JsDESL}.

@@

\subsection{LMS technical overview}

In LMS, a DSL is split into two parts, its
interface and its implementation. Both parts can be assembled from
components in the form of Scala traits. DSL programs are written in
terms of the DSL interface only, without knowledge of the
implementation.

Part of each DSL interface is an abstract type constructor
\code{Rep[_]} that is used to wrap types in the DSL programs. The DSL
implementation provides a concrete instantiation of \code{Rep} as IR
nodes. When the DSL program is staged, it produces an intermediate
representation (IR), from which the final code can be generated. In
the DSL program, wrapped types such as \code{Rep[Int]} represent
staged computations while expressions of plain unwrapped types
(\code{Int}, \code{Bool}, etc.) are evaluated at staging time as
in~\cite{finally-tagless,polymorphic-embedding}.

Consider the difference between these two programs:
\begin{lstlisting}
def prog1(b: Bool, x: Rep[Int]) = if (b) x else x+1
def prog2(b: Rep[Bool], x: Rep[Int]) = if (b) x else x+1
\end{lstlisting}

The only difference in these two programs is the type of the parameter
\code{b}, illustrating that staging is purely type-driven with no
syntactic overhead as the body of the programs are identical.

In \code{prog1}, \code{b} is a simple boolean, so it must be provided
at staging time, and the \code{if} is evaluated at staging time. For
example, \code{prog1(true, x)} evaluates to \code{x}. In \code{prog2},
\code{b} is a staged value, representing a computation which yields a
boolean. So \code{prog2(b, x)} evaluates to an IR node for the \code{if}:
{\tt\small If(b, x, Plus(x, Const(1)))}.

For \code{prog2}, notice that the \code{if} got transformed into an IR
node. To achieve this, LMS uses
Scala-Virtualized~\cite{scala-virtualized}, a suite of minimal
extensions to the regular Scala compiler, in which control
structures such as \code{if} can be reified into method calls, so that
alternative implementations can be provided. In our case, we provide an
implementation of \code{if} that constructs an IR node instead of
acting as a conditional. In addition, the \code{+} operation is
overloaded to act on both staged and unstaged expressions. This is
achieved by an implicit conversion from \code{Rep[Int]} to a class
\code{IntOps}, which defines a \code{+} method that creates an IR node
\code{Plus} when executed. Both of \code{Plus}'s arguments must be
staged. We use an implicit conversion to stage constants when needed
by creating a \code{Const} IR node.

\subsection{Example: a DSL program and its generated JavaScript code}

The following DSL snippet creates an array representing a table of
multiplications:
\begin{lstlisting}
def test(n: Rep[Int]): Rep[Array[Int]] =
  for (i <- range(0, n); j <- range(0, n)) yield i*j
\end{lstlisting}

Here is the JavaScript code generated for this snippet:
\begin{lstlisting}[language=JavaScript]
function test(x0) {
  var x6 = []
  for(var x1=0;x1<x0;x1++){
    var x4 = []
    for(var x2=0;x2<x0;x2++){
      var x3 = x1 * x2
      x4[x2]=x3
    }
    x6.splice.apply(x6, [x6.length,0].concat(x4))
  }
  return x6
}
\end{lstlisting}

The generated code resembles single-assignment form. The nested
\code{for}-loop is desugared into a \code{flatMap} which generates the
nested \code{for}-loop and the \code{splice} pattern concatenating the
inner \code{x4} arrays into one \code{x6} array in the JavaScript
code.\footnote{Obviously, the generated code can be optimized further.}

\subsection{Walkthrough: defining a DSL component}\label{sec:defineDslComponent}

To conclude the introduction to LMS, we show how to add a component
for logging in a DSL, generating JavaScript code which calls
\code{console.log}.

We start by defining the interface:
\begin{lstlisting}
trait Debug extends Base {
  def log(msg: Rep[String]): Rep[Unit]
}
\end{lstlisting}

The \code{Base} trait is part of the core LMS framework and provides
the abstract type constructor \code{Rep}.

Now, we define the implementation:
\begin{lstlisting}
trait DebugExp extends Debug with EffectExp {
  case class Log(msg: Exp[String]) extends Def[Unit]
  def log(msg: Exp[String]): Exp[Unit] = reflectEffect(Log(msg))
}
\end{lstlisting}

The \code{EffectExp} trait is part of the core LMS framework. It
inherits from \code{BaseExp} which instantiates \code{Rep} as
\code{Exp}. \code{Exp} represents an IR via two subclasses:
\code{Const} for constants and \code{Sym} for named values defining a
\code{Def}. \code{Def} is the base class for all IR nodes. In our
\code{DebugExp} trait, we extend \code{Def} to support a new IR node:
\code{Log}.

IR nodes are defined as \code{Def}s but they are never referenced
explicitly as such. Instead each \code{Def} has a corresponding symbol
(an instance of \code{Sym}). IR nodes refer to each other using their
symbols. This is why, in the code shown, the \code{msg} parameter is
of type \code{Exp} (not \code{Def}). The method \code{log} 
returns an \code{Exp}. Calling \code{reflectEffect} is what creates
this symbol from the \code{Def}.

In general, the framework provides an implicit conversion from
\code{Def} to \code{Exp}, which performs common subexpression
elimination by re-using the same symbol for identical definitions. We
do not use the automatic conversion here, because \code{log} is a 
side-effecting operation, and we do not want 
to (re)move any such calls even if their message
is the same.

The framework schedules the code generation from the graph of
\code{Exp}s and their dependencies through \code{Def}s. It chooses
which \code{Sym}/\code{Def} pairs to emit and in which order. To
implement code generation to JavaScript for our logging IR node, we
simply override \code{emitNode} to handle \code{Log}:
\begin{lstlisting}
trait JSGenDebug extends JSGenEffect {
  val IR: DebugExp
  import IR._
  override def emitNode(sym: Sym[Any], rhs: Def[Any])(
    implicit stream: PrintWriter) = rhs match {
      case Log(s) => emitValDef(sym, "console.log(" + quote(s) + ")")
      case _ => super.emitNode(sym, rhs)
    }
}
\end{lstlisting}
Notice that in order to compose nicely with other traits, the
overridden method just handles the case it knows and delegates to
other traits, via \code{super}, the emitting of nodes it doesn't know
about.

@@

In a Web application, the user interface is defined by a HTML document that can be updated by the
JavaScript code. A typical operation consists in searching some "interesting" element in the
document, in order to extract its content, replace it or listen to user events triggered on it (such
as mouse clicks). The standard API provides several functions to search elements in a HTML document
according to their name or attribute values. Figure~\ref{selectors-api} summarizes the available
functions and their differences.

@@

\bibliographystyle{splncs}
%\bibliographystyle{plain}
\bibliography{biblio}

\end{document}
